---
title: "Session 1: Introduction and Visualisation Extended (Template)"
subtitle: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    includes:
    code_folding: "show"
    toc: true
    toc_float: true
    number_sections: false
---

# Setup 


### Start up  - Folder structure
(same as previous Rmd no need to do it twice)

### Download datasets 

- ExerciceBisphenol.csv 
- ExercicePlasmaCholesterol.xlsx

And save them into your "data" folder.


### Same .Rproject 

Double check that you are still in the same Rproject (check on the top right of RStudio)


### Load libraries

```{r, message=FALSE, warning=FALSE}
#load required packages
library(ggplot2) #to plot
library(readr) #to read files
library(here) #to find files


#if you want to make it reproducible (more advance)
# if(!require(pacman)) {install.packages(c("pacman"))}
# 
# pacman::p_load(here,ggplot2, readr)


#these are just Rmarkdown visual options
knitr::opts_chunk$set(message=F, warning=F, comment=NA) 


```



Find your path

```{r}
here() #from the here packages is a neat magick trick that will find where you files are 
#given that you are in a .Rproject, type help(here) for more info

```

# Read the data

```{r}

il18 <- read_csv(here("data", "freeIL18.csv")) #read the data
#you can also just read it manually
#il18 <- read_csv("C:/Users/david/Documents/cours_stat/freeIL18.csv")
il18


names(il18) <-  c("samples", "dis.act", "freeIL18", "TotalIL18") # assign new names to columns


```

# Visualizations

## Numerical

### Density Plots


Show the density of a numeric variable

A density plot shows the distribution of a numeric variable. It takes only numeric variables as input and is very close from an histogram. It can be use in the same exact condition.

Tips:

-   Play with the bandwidth of your density function.

-   Don't show the distribution of more than \~5 variables. Use Violin or Ridge line plot instead.

-   Avoid filling with color palettes.

#### Do a Basic density plot: x = freeIL18

```{r}


#hint: ggplot(data, aes(...)) + geom_density()
#hint2: assign your plot to an object, e.g. density <- ggplot(data, aes(...)) + geom_density()



```

####  Density plot by groups (dis.act)

```{r}

#hint use fill


```

####  Stacked density plot

```{r}

#hint: same as before but use position="fill" in geom_density


```



#  Data Manipulation

### Load libraries

```{r, message=FALSE, warning=FALSE}

#install.packages("tidyr") #  data manipulation
#install.packages("readxl") # read excel files
#install.packages("writexl") # write excel files
#install.packages("scales") # work with different scales
 #<- you just need to run these lines once


library(tidyr)
library(readxl)
library(writexl)
library(scales)

```


#### Try to read a .csv file: ExerciceBisphenol.csv
```{r}

#hint: use read_csv() and call it df_bisphenol


```


#### Try to read an excel file

```{r}

#hint: use read_xlsx() and call it plasma


```

#### Look at the data


Output the summary of the Bisphenol data

```{r}
#hint: use summary() and or str()

```

Same for the Plasma data

```{r}
#hint use summary() and or str()

```



### Let's look deeper into plasma

We will start with plasma. We have names with spaces, let's rename them !

```{r}
#rename the columns of plasma data to "control","treatment_1","treatment_2"
#hint: use names() and a vector of names i.e. c("..",  "..")


```


#### Wide to long format
Wide, or unstacked data is presented with each different data variable in a separate column.

Long, or narrow data is presented with one column containing all the values and another column listing the context of the values.

Many functions in R expect data to be in a long format rather than a wide format.

Here the data is in a wide format, let's put it in long format so we can use all the power of mist R functions.

For that we use the package **tidyr**

```{r}

plasma_long <- pivot_longer(data = plasma, c("control","treatment_1","treatment_2"))
plasma_long
```
Reflect: Do you see the differences ?


#### Now do a boxplot : x = name, y = value and add point

```{r}

#hint: use geom_boxplot + geom_jitter
#hint2: you might want to rename the column ?
#bonus: try to do the same plot with plasma (the one in wide format)
 



```

### Let's look deeper into  bisphenol



Contingency tables of some experiment, to check is groups are of the same length.


```{r}
df_bisphenol$tot <- df_bisphenol$Cancer + df_bisphenol$NoCancer
df_bisphenol
```


#### Let's try a barplot with this wide format data:

First just plot the "cancer"

NB: we use geom_col

```{r}
wide_plot <- ggplot(df_bisphenol)+
  geom_col(aes(group,Cancer,fill = "cancer"))

wide_plot
```

Then add the "no cancer"

```{r}
wide_plot+
  geom_col(aes(group,NoCancer,fill = "no cancer"))
```
#### Problem! 

The second bar is over the first one! Need to be in **long** format too! Can you do it ?

```{r}

#hint: use pivot_longer


```

#### Change from "stacked" plot to "dodged" plot
```{r}

#hint: do the same thing but add position = "dodge" to geom_col()

```


#### Add percents
```{r}

#hint: do the same thing as step one but add scale_y_continuous(labels = percent_format(accuracy = 1)) at the end


```


# More Visualisation Stuff

### Use themes

##### Default

```{r}

violin <- ggplot(data = il18, aes(x = dis.act, y = freeIL18, fill=dis.act )) +
  geom_violin() + 
  xlab("Disease activity")

violin + theme_gray() # the default
```

##### Black and White

```{r}
violin + theme_bw() # the default
```

##### Classic

```{r}
violin + theme_classic() # the default
```



### Interactive plots

The package **plotly** helps creating awesome interactive data vizualisation and maks it super easy to do !

First load it

```{r}

#install.packages(plotly) # install it if you don't have it

library(plotly)
```

Lets redo the last scatter plot but this time with a twist !

```{r}
plot_object <- ggplot() + 
  geom_point(data = il18,
             aes(x = freeIL18, y = TotalIL18,  color=dis.act)) +
  geom_point()

ggplotly(plot_object)
```
