---
title: "session 3 visualization tests"
output:
  html_document:
    df_print: paged
---

Let's load the different examples files. Replace the file path `C:/Users/dmog/Documents/cours/stat_2022/Travaux_Pratiques/Session4_tets_partie2/data/` with yours.

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggbeeswarm)

df_bisphenol <- read.csv("C:/Users/dmog/Documents/cours/stat_2022/Travaux_Pratiques/Session4_tets_partie2/data/ExerciceBisphenol.csv")

plasma <- readxl::read_xlsx("C:/Users/dmog/Documents/cours/stat_2022/Travaux_Pratiques/Session4_tets_partie2/data/ExercicePlasmaCholesterol.xlsx")

pairedData <- readxl::read_xlsx("C:/Users/dmog/Documents/cours/stat_2022/Travaux_Pratiques/Session4_tets_partie2/data/pairedData.xlsx")


```

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = F)
```


# The functions to use:
## continuous variables

T test
```{r,eval=F}
t.test
```

Wilcoxon (paired or not)

```{r,eval=F}
wilcox.test
```

ANOVA (analysis of variance)

```{r,eval=F}
aov
```

Kruskal Wallis

```{r,eval=F}
kruskal.test
```


## binary variables

Fisher

```{r,eval=F}
fisher.test
```

Mc Nemar
```{r,eval=F}
mcnemar.test
```

Chi 2
```{r,eval=F}
chisq.test
```


# Plasma
## Try to tranform to long:

```{r}
plasma
```

```{r}
ggplot(plasma)+
  geom_boxplot(aes(x = "control",y = Control))+
  geom_boxplot(aes(x = "treatment 1",`Treatment 1`))+
  geom_boxplot(aes(x = "treatment 2",`Treatment 2`))
```



```{r}
plasma_long <- pivot_longer(plasma,
                            cols = c("Control","Treatment 1","Treatment 2"),
                            values_to = "plasma",
                            names_to = "group")
plasma_long
```

## make a plot (same as session 2)


```{r}
ggplot(data=plasma_long,
       aes(x = group,
           y = plasma,
           fill=group))+  #define data and variables
  geom_boxplot()+
  geom_jitter(shape=21,show.legend = F,cex=3) #create one layer of graph

```

## test assumptions

- gaussian distrib in each group

```{r}
hist(plasma$Control)
```

```{r}
hist(plasma$`Treatment 1`)
```

```{r}
hist(plasma$`Treatment 2`)
```

```{r}
plasma_long %>%
  ggplot()+
  geom_histogram(aes(plasma),
                 binwidth = 5,
                 color = "black")+
  facet_wrap(~group,scales = "free")
```

- variance similar

base:

```{r}
var(plasma$Control)
var(plasma$`Treatment 1`)
var(plasma$`Treatment 2`)
```

Attention, la vairable treatment 2 contient des manquants!

```{r}
plasma$`Treatment 2`
```

Il faut spécifiquement dire que il y a des manquants:

```{r}
var(plasma$`Treatment 2`,na.rm = T)
```


dplyr like:

```{r}
plasma_long %>%
  group_by(group)%>%
  dplyr::summarise(mean = mean(plasma,na.rm = T),
                   sd = sd(plasma,na.rm = T),
                   var = var(plasma,na.rm = T))
```


Perform Anova and/or Kruskal-Wallis

```{r}
aov_res <- aov(plasma ~ group,
               data = plasma_long)

summary(aov_res)
```

- what does it means ? What is F ? What is p ?

F value is the value of the test statistics (ratio of inter group variance and intra group variance). The test statistics follow a F distribution

```{r}
N <- nrow(na.omit(plasma))
x <- seq(0,10,0.1)
data.frame(x = x,
           df = df(x,N,N),
           pf = pf(x,N,N)) %>%
  ggplot()+
  geom_line(aes(x,df,color = "df"))+
  geom_line(aes(x,pf,color = "pf"))+
  geom_hline(yintercept = .95)+
  geom_vline(xintercept = 2.2)+
  theme_bw()+theme(legend.position = c(1,1),
                   legend.justification = c(1,1))+
  labs(color = "")
```

Values of F above 2.2, i.e. when inter group variance is at least 2.2 intra group variance, implies that there is less than 5% of chance to obtain such test statistics or higher under the null Hypothesis.


```{r}
krus_res <- kruskal.test(plasma ~ group,data = plasma_long)
krus_res
```

- make Tukey pairwise comparison

```{r}
TukeyHSD(aov_res)
```



# paired data

- The level of a marker is measured at baseline and 1 week after injection of a product
- Sample: N=20 rabbits
- Aim: to assess and test the effect of the product between baseline and 1 week


```{r}
pairedData
```

## verify assumptions

- gaussian ?
- number of observations ?

```{r}
hist(pairedData$T0)
```

```{r}
hist(pairedData$W1)
```

Not Gaussian! 

Other way: long format, then ggplot

```{r}
paired_long <- pairedData %>% 
  pivot_longer(cols = c("T0","W1"),
               names_to = "time")
```


```{r}
paired_long %>%
  ggplot(aes(value))+
  geom_histogram(binwidth = 3,color = "black")+
  facet_wrap(~time)
```


## do the tests

- perform the proper test

The proper test is 

```{r}
wilcox.test(pairedData$T0,pairedData$W1,paired=T)
```

```{r}
wilcox.test(value ~ time, data = paired_long)
```

- see the difference if using not paired 

```{r}
wilcox.test(pairedData$T0,pairedData$W1,paired=F)
```

explanation:

If one looks without pairing, group do not look different:

```{r}
pairedData %>% 
  pivot_longer(cols = c("T0","W1"),
               names_to = "time")%>%
  ggplot(aes(time,value))+
  geom_jitter()+
  geom_boxplot(alpha = .3,outlier.shape = NA)
```

But if we connect lines, we see that most of the measures increase

```{r}

paired_long$timejitter <- jitter(as.numeric(factor(paired_long$time)))

paired_long %>%
  ggplot(aes(timejitter,value))+
  geom_point(aes(color = time))+
  geom_line(aes(group = id))+
  scale_x_continuous(breaks = c(1,2),labels = c("T0","W1"))
```

An easy way to see it: look at the paired differences:

```{r}
pairedData %>%
  ggplot(aes(x = 1,y = W1-T0))+
  geom_jitter()+
  geom_boxplot(alpha = .2)
```



```{r}
t.test(pairedData$T0,pairedData$W1,paired=T)
```


## see test change

See if the ratio of level > 1 changes between baseline and 1 W



Let's test if the rabbits who were above 1 at T0 are still above 1 at W1

```{r}

concordanceT0W1 <- table(pairedData$T0>1,pairedData$W1>1,dnn = c("T0","W1"))
concordanceT0W1

```



```{r}
mcnemar.test(concordanceT0W1)
```



