---
title: "session 4  tests"
output:
  html_document:
    df_print: paged
---

Let's load the different examples files. Replace the file path `C:/Users/dmog/Documents/cours_stat/Data-20220301/` with yours.

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggbeeswarm)

df_bisphenol <- read.csv("C:/Users/dmog/Documents/cours/stat_2022/Travaux_Pratiques/Session4_tets_partie2/data/ExerciceBisphenol.csv")

plasma <- readxl::read_xlsx("C:/Users/dmog/Documents/cours/stat_2022/Travaux_Pratiques/Session4_tets_partie2/data/ExercicePlasmaCholesterol.xlsx")

pairedData <- readxl::read_xlsx("C:/Users/dmog/Documents/cours/stat_2022/Travaux_Pratiques/Session4_tets_partie2/data/pairedData.xlsx")



```


# The function to use:
## continuous variables

T test
```{r}
t.test
```

Wilcoxon (paired or not)

```{r}
wilcox.test
```

ANOVA (analysis of variance)

```{r}
aov
```

Kruskal Wallis

```{r}
kruskal.test
```




## binary variables

Fisher

```{r}
fisher.test
```

Mc Nemar
```{r}
mcnemar.test
```

Chi 2
```{r}
chisq.test
```


# Plasma data

Same data as last session. Now we want to test if there is a difference between the three groups!

```{r}
plasma
```
## Try to tranform to long

Have a look at last session

## make a plot (same as session 2)


## test assumptions

- Gaussian distribution for each group
- similar variance

## do the test

- Perform Anova and/or Kruskal-Wallis
- what does it means ? What is F ? What is p ?
- perform Tukey pairwise comparison, using the function `TukeyHSD`


# paired data

- The level of a marker in blood is measured at baseline (T0) and 1 week (W1) after injection of a product
- Sample: N=20 rabbits
- Aim: to assess and test the difference in marker between baseline and 1 week

```{r}
pairedData
```


## verify assumptions

- Gaussian ?
- number of observations ?

## do the tests

- perform the proper test
- see the difference if using not paired 

## see test change

See if the proportion of the blood marker levels > 1 versus < 1 changes between baseline and W1. Which test should you use to show significance ? Perform the test.


