---
title: "Session R: Correlation and Regression"
subtitle: "Denis Mongin et David Munoz Tord"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    includes:
    code_folding: "show"
    toc: true
    toc_float: true
    number_sections: false
---


# Setup



## Start up  - Folder structure

Create a folder

Inside the folder named "Stats_2022" (or whatever you named it)  create a folder named "Session_6" (or whatever similar to your liking)

And inside this one create a folder names "data" (here I really suggest you call it "data")

### Download dataset

- session6_diet.xls
- session6_eggs.xls

And save it into your "data" folder.


### Create an .Rproject 

Within RStudio go into Files -> New Project

Choose: Existing Project and browser the "Session_6" folder you just created (or whatever you named it)

This will reload RStudio, don't worry


### Load libraries

```{r, message=FALSE, warning=FALSE}
#load required packages
library(dplyr) # to wrangle dats
library(tidyr) # to tidy the data
library(readxl) # to read the files
library(here) # to find files
library(ggplot2)


#these are just Rmarkdown visual options
knitr::opts_chunk$set(message=F, warning=F, comment=NA) 


```

Know where you are !

```{r}
#here() #from the here packages is a neat magic trick that will find where you files are 
#given that you are in a .Rproject, type help(here) for more info

```

## Read the data

```{r}

df_diet <- read_xls(here("data", "session6_diet.xls")) #read the data
#you can also just read it manually
#il18 <- read_xls("C:/Users/david/Documents/cours_stat/session6_diet.xls") # but this get annoying and buggy very fast
df_eggs <- read_xls(here("data", "session6_eggs.xls"))


```


Eggs is for the Platorchestia platensis example, and diet is the weight loss for subject ongoing a diet. 

We will use the eggs data for correlations, and the diet one for linear regression.


# Correlation with the Eggs Data


Check out the column names and data

```{r}
head(df_eggs)
```

##  Visulize the data

Make a plot of weigh in function of number of eggs:


```{r}

#hint: ggplot(data, aes(...)) + geom_point()
#hint2: assign your plot to an object, e.g. scatter <- ggplot(data, aes(...)) + geom_point()

scatter <- ggplot(data = df_eggs, aes(x = NbEggs, y = weight)) +
  geom_point()

scatter 

```

Inspect the data, is there anything that you should look for when doing correlations ? Is there any "problem" or can we continue ?



## Test the Correlation

Now estimate the correlation between the number of eggs and their weight

```{r}
#hint: use cor.test(data$x, data$y)

cor.test(df_eggs$weight, df_eggs$NbEggs)

```
Now run this chunk and read a bit the cor.test documentation

```{r}

?cor.test

```


Finally try to compute the Pearson and Spearman correlation of the same previous variables

```{r}
#hint: use cor.test(data$x, data$y, method = "..") 
cor.test(df_eggs$weight, df_eggs$NbEggs, method = "spearman")
```

Try to explain the different results with your neighbor.

## Linear regression with the Diet Data 



Check out the column names and data

```{r}
head(df_diet)
```

Now were are going to:


- Do a linear regression predicting the number of eggs in function of the weight using `lm()` 
- Inspect the coefficient using `summary()`
- Try to understand what does the coefficient mean ? 
- Predict the Number of eggs of a 15 g Platorchestia 



## Plot



represent the relationship between the weight loss `Weight.Diff` and `Age`, `height`, and `pre.weight`

## correlations

test the correlations between `Weight.Diff` and the other variables

## regression

- perform the regression `Weight.Diff` in function of the weight before pre.weight, and interpret the result.





