---
title: "Session 1: Homework"
subtitle: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    includes:
    code_folding: "show"
    toc: true
    toc_float: true
    number_sections: false
---

### Download the dataset on moodle (folder Data)

- data_RT.csv

And save it into your "data" folder.


### Load libraries

```{r, message=FALSE, warning=FALSE}
#load required packages
library(ggplot2) #to plot
library(readr) #to read files
library(here) #to manages paths

```


### Read the data and check it out

A little bit about the data.

This is a modified dataset from a cognitive neuroscience experiment.

It basically tried to compare typical adults (TA) vs patients with Alzheimer's disease (AD) on a reaction time (RT) task. 

Participants also had to go through a cognitive intervention or went through a control condition

So you have

- id: Each participants experiment ID
- intervention: Wether or not the participant went trough the intervention or the control condition
- type: Wether the participant was disgnosed with AD or not
- RT: Each participant's reaction time to a cognitive task


```{r}

#hint: use read_csv or double click on the file and import
#hint2: watch out what name is assigned to the data !

data_RT  <- read_csv("C:/Users/dmunoz/Downloads/data_RT.csv") #read the data
#you can also just read it manually
#daat_RT <- read_csv("C:/Users/david/Documents/cours_stat/data_RT.csv")
data_RT


```



### Do a stacked barplot: 


```{r}

#hint: x = intervention and fill with type

barplot <- ggplot(data_RT, aes(intervention, fill=type)) +  #here we only need on variable so ggplot will automatically assign it to
# to the x-axis
 geom_bar()

barplot

```

### Do a basic histogram: 

```{r}

#hint: x = RT


hist <- ggplot(data_RT, aes(RT)) + 
  geom_histogram()

hist

```

### You are all done !