---
title: "Session 1: Homework"
subtitle: "Your Name"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    includes:
    code_folding: "show"
    toc: true
    toc_float: true
    number_sections: false
---

### Download the dataset on moodle (folder Data)

-   data_RT.csv

And save it into your "data" folder.

### Load libraries

```{r, message=FALSE, warning=FALSE}
#load required packages
library(ggplot2) #to plot
library(readr) #to read files
library(here) #to manages paths

```

### Read the data and check it out

A little bit about the data.

This is a modified dataset from a cognitive neuroscience experiment.

It basically tried to compare typical adults (TA) vs patients with Alzheimer's disease (AD) on a reaction time (RT) task.

Participants also had to go through a cognitive intervention or went through a control condition

So you have

-   id: Each participants experiment ID
-   intervention: Whether or not the participant went through the intervention or the control condition
-   type: Whether the participant was diagnosed with AD or not
-   RT: Each participant's reaction time to a cognitive task

```{r}

#hint: use read_csv or double click on the file and import
#hint2: watch out what name is assigned to the data !
df <- read_csv('data/data_RT.csv')
df
```

### Do a stacked barplot:

```{r}

#hint: x = intervention and fill with type
ggplot(data=df, aes(intervention, fill=type)) +
geom_bar()
```

### Do a basic histogram, then add a vertical line for the median and another for the mean values:

```{r}

#hint: x = RT
ggplot(data=df, aes(RT)) + 
geom_histogram() + 
geom_vline(xintercept = mean(df$RT), col="green") +
geom_vline(xintercept = median(df$RT), col="blue")

```

### You are all done !
