---
title: "Session 2: Parameters Estimate"
subtitle: "Your Name"
date: "October 16, 2022"
output:
  html_document:
    includes:
    code_folding: "show"
    toc: true
    toc_float: true
    number_sections: false
---

# Setup

```{r setup, include=FALSE}

knitr::opts_chunk$set(message=F, warning=F, comment=NA) #these are just Rmarkdown visual options

```

### Load libraries

```{r}

library(ggplot2)

```

# Simulation

The goal of this code is to create a small simulation study to:

1.  illustrate central limit theorem

2.  show that 95% of generated samples should contain the true parameters

### Let's generate some data

```{r}

#random generation for the normal distribution with mean equal to mean and standard deviation equal to sd.
population <- rnorm(10000, mean=0, sd=1)

ggplot() + geom_histogram(aes(population)) #do a simple histogram
```

We now generate a population of 1000 individuals following a Gaussian distribution.

### Resample

We will now randomly sample 50 individuals from this population.

```{r}
sample_size <- 10 # you can play around with this number

mysample <- sample(population, sample_size) #take a data + a sample size 

ggplot() + geom_histogram(aes(mysample)) #do a simple histogram

```

Can you describe the differences ?

Play around with the sample size and see what happens. (Just change "50" to something else)

### Calculate sample's statistics

```{r}
summary(mysample)
```

We now create 100 sample of size `sample_size`, with their statistics:

```{r}

size = 100

#this creates a dataframe with fix parameters (do not need to understand this code)
#rep() juste repeat a number over the number of time you want

# this is are all random number has placeholder for later
mySampleRes <- data.frame(mean=rep(0,size),
                          sd=rep(1,size),
                          sample = rep(NA,size),
                          se=rep(1/sqrt(10),size), 
                          confintLow=rep(0,size),
                          confintHi=rep(0,size))

head(mySampleRes) # all rows are the same for now
```

the real mean of my original population is 0


```{r}

#loop through each row

for (i in 1:size){
  mysample <- sample(population, sample_size)#sample from population, same as above
  mySampleRes$sample[i] <- i #assign an incemental index 
  mySampleRes$mean[i] <- mean(mysample) #calculate the mean of this random sample
  mySampleRes$sd[i] <- sd(mysample) #calculate the standard deviation of this random sample
  mySampleRes$min[i] <- min(mysample) #calculate the minimum of this random
  mySampleRes$se[i] <- mySampleRes$sd[i]/sqrt(sample_size) #calculate the sampling error of this random sample
  mySampleRes$confintLow[i] <- mySampleRes$mean[i]-1.96*mySampleRes$se[i] #calculate the lower 95% confidence interval of this random sample
  mySampleRes$confintHi[i] <- mySampleRes$mean[i]+1.96*mySampleRes$se[i] #calculate the upper 95% confidence 
}

head(mySampleRes)


# for reference:
# Confidence level Confidence interval (mean ± sampling error)
# 68% mean ± (1.0) x (SE)
# 95% mean ± (1.96) x (SE)
# 99% mean ± (2.58) x (SE) 
# Bonus: Can you describe/explain/visualize why ?
```

```{r}

mySampleRes$within_confint <- factor(mySampleRes$confintLow < 0 & mySampleRes$confintHi > 0, labels = c("Outside", "Inside")) # create variable to check if we have found and estimate (mean) which its confindence intervals doesn't include 0


table(mySampleRes$within_confint)

```




Let's plot it !

```{r}
ggplot(mySampleRes,
       aes(sample,mean,color = within_confint))+
  geom_point()+
  geom_errorbar(aes(ymin = confintLow,ymax = confintHi))+
  geom_hline(yintercept = 0,color = "green")+
  coord_flip()+
  labs(title = "Mean of 100 samples")
```

Red estimates are different than zero with 95% confidence

### Distribution of the means

```{r}
p_sample <- ggplot()+
  geom_histogram(data = mySampleRes,
                 aes(x=mean),binwidth = .2) +
  xlim(round(min(population)),round(max(population))) +
  geom_vline(xintercept = 0,color = "green") +
  labs(title = "Sampled Means")

p_sample
```

Distribution of the means depends on the sample size.

```{r}
library(patchwork)
#install.packages('patchwork')
p_population <- ggplot()+
                geom_histogram(data = data.frame(x = population),
                               aes(x=x),binwidth = .2)+
                  xlim(round(min(population)),round(max(population)))+
                  labs(title = "population")


p_population / p_sample
```


# same with the min

What we did with the mea cannot work with the min, or max, or sd.
Let's see. The true min is:

```{r}
true_min <- min(population)
```

I create a confidence interval of teh sampled min:

```{r}

mySampleRes$confintLow_min <- mySampleRes$min-1.96*mySampleRes$se #calculate the lower 95% confidence interval of this random sample
mySampleRes$confintHi_min <- mySampleRes$min+1.96*mySampleRes$se #calculate the upper 95% confidence interval of this random sample
```

I create a variable saying if the true value is within my confidence interval:
```{r}

mySampleRes$within_confint_min <- ifelse(mySampleRes$confintLow_min < true_min & mySampleRes$confintHi_min > true_min,"Outside","Inside") # create variable to check if we have found and estimate (mean) which its confindence intervals doesn't include 0
  
```


```{r}
ggplot(mySampleRes,
       aes(sample,min,color = within_confint_min))+
  geom_point()+
  geom_errorbar(aes(ymin = confintLow_min,ymax = confintHi_min))+
  geom_hline(yintercept = true_min,color = "green")+
  coord_flip()+
  labs(title = "MIn of 100 samples")
```

We see most of sampled min are too far fro the true min! The distribution of the minimum is not centered on the true min:


```{r}
p_sample_min <- ggplot()+
  geom_histogram(data = mySampleRes,
                 aes(x=min),
                 binwidth = .2) +
  xlim(round(min(population))-1,round(max(population))) +
  geom_vline(xintercept = true_min,color = "green") +
  labs(title = "Sampled Means")

p_sample_min
```

