---
title: "Session 6"
output: html_notebook
---
This are the exercise for session 6.

We will use two datasets: session6_diet.xls and session6_eggs.xls

```{r}
library(dplyr)
library(tidyr)
library(readxl)

df_diet <- read_xls("C:/Users/dmog/Documents/cours_stat/Data-20220301/session6_diet.xls")
df_eggs <- read_xls("C:/Users/dmog/Documents/cours_stat/Data-20220301/session6_eggs.xls")
```

Eggs is for the Platorchestia platensis example, and diet is the weight loss for subject ongoing a diet. We will the eggs data for correlations, and the diet for linear regression

# eggs and correlation

check what is inside:


```{r}
df_eggs
```

##  plot

- make a plot of weigh function of eggs

## correlation

- estimate the correlation with 

```{r}
cor.test()
```

- do ?cor.test and try to do Pearson and Spearman. Explain the difference

## linear regression

- do a linear regression predicting the number of eggs in funciton of the weight using `lm()` 
- inspect the coefficient using `summary()`
- What do the coefficient mean ? 
- could you predict the Number of eggs of a 15 g Platorchestia ?

# diet

```{r}
df_diet
```

## plot

represent the relationship between the weight loss `Weight.Diff` and `Age`, `height`, and `pre.weight`

## correlations

test the correlations between `Weight.Diff` and the other variables

## regression

- perform the regression `Weight.Diff` in function of the weight before pre.weight, and interpret the result.





